---
title: "README"
author: "Daniela Parada"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
   md_document:
   variant: markdown_github
---

The following real data example of the implementation of robust estimators for functional quadratic regression models. These robust estimators  of the principal directions with robust regression estimators based on a bounded loss function and a preliminary residual scale estimator. This is part of a work in progress done in collaboration with Prof. Dr. Graciela Boente.

Let's first load some <code>R</code> packages.

```{r installation, results='hide', cache=FALSE, results=FALSE, warning=FALSE, comment=FALSE, message=FALSE}
library('fda')          
library('robustbase')   # to use lmrob 
library(fda.usc)        # tecator dataset and fda tools
library(gdata)          # to use uppertriangle
library(lattice)        # to plot
```

Let's now load some custom <code>R</code> functions that are needed to compute robust estimators.

```{r installationcustom, results='hide', cache=FALSE, results=FALSE, warning=FALSE, comment=FALSE, message=FALSE}
source('funciones-auxiliares.R') 
source('descomponer3.R')         # for covariance decomposition
source('minimizar.R')            # minimization menu
```


We will use the Tecator data set available in <code>R</code> (http://lib.stat.cmu.edu/datasets/tecator). Each observation consists of a spectrometric curve that corresponds to the absorbance measured on an equally spaced grid of 100 wavelengths between 850 and 1050 nm. The contents of fat protein and moisture were also recorded through analytic chemistry methods. The goal of the analysis is to predict the fat content ($y$) using some characteristics of the spectrometric curve. 

```{r loading data}
datos <- data(tecator)
absorp <- tecator$absorp.fdata
absorp1 <- fdata.deriv(absorp,nderiv = 1)  #  computes the first derivative
```

Plots of the spectrometric data and it's first derivative are displayed below.

```{r initial plots, echo=FALSE, fig.height=6, fig.width=16}
par(mfrow=c(1,2))
plot(absorp)  #  spectrometric curve
plot(absorp1, main="First derivative")  #  first derivative 
par(mfrow=c(1,1))
```


Functional boxplots of the spectrometric data and it's first derivative are displayed below.

```{r functional boxplots, echo=FALSE, fig.height=6, fig.width=16}
par(mfrow=c(1,2))
# First derivative
fd1 <- fbplot(t(absorp1$data), x = absorp1$argvals, method =  "MBD",
       xlab = "Wavelength", ylab = "d(Absorbance, 1)",
       xlim = c(850, 1050), ylim = c(-0.02, 0.05))

# Absorbances
fd0<-fbplot(t(absorp$data), x = absorp$argvals, method = "MBD",
       xlab = "Wavelength", ylab = "Absorbance",
       xlim = c(850, 1050), ylim = c(2, 5.5))
par(mfrow=c(1,1))
```


In particular, we will now consider the first derivative of the spectrometric curve which we denote $X$.

```{r X(t), echo=FALSE, fig.width=16, fig.height=6}
matplot(absorp1$argvals, t(absorp1$data), type = "l", lty = 1,
        xlab = "Wavelength (nm)", ylab = "X(t)") 
```


# Robust Estimators under a Functional Linear/Quadratic Model

FLM: $$y=\alpha_0+\langle X, \beta_0\rangle + \epsilon$$

FQM: $$y=\alpha_0+\langle X, \beta_0\rangle +\langle X, \Upsilon_0 X\rangle+ \epsilon$$

In both cases, we choose 4 principal directions which explain more than 98% of the total variability as seen below.

```{r robust est, echo=FALSE}
set.seed(124) 
covariable <- 'd1' # first derivative  
 
indices <- 1:215  

# indices
tecdatos <- submuestra(indices, covariable = covariable)
 
indices_tecdatos <- indices
dt_tecdatos <- tecdatos$t[2]-tecdatos$t[1]



###############################################
## Choose the number of principal directions
## equal to 4, then varexp should be 1
###############################################


freq=4
varexp=1

est_rob_LINEAL <- estimar(y=tecdatos$y,xcenter= tecdatos$x,  
			ttt= tecdatos$t, ajuste='lineal', freq=freq,  cov_type='gerS',  
			fLoss='lmrob',cterho=3.443689, nresamp=5000, varexp=varexp) 

#######################################
# PERCENTAGE EXPLAINED By 4 directions
#######################################

est_rob_LINEAL$porcentaje

#0.9841754

#############################################
# Store the estimates and compute predictions
# for the residuals
#############################################

beta_rob_LINEAL <- est_rob_LINEAL$beta
 
phies_rob_LINEAL <- est_rob_LINEAL$autofun

coef_rob_LINEAL <-   tecdatos$x %*% phies_rob_LINEAL * dt_tecdatos

predichos_ROB_LINEAL <- est_rob_LINEAL$alfa + coef_rob_LINEAL%*% est_rob_LINEAL$slope_coef  


#####################################################################################
## ROBUST ESTIMATOR UNDER A FQM 
## fat= alfa_0 + < absorp1, beta_0 > + <absorp1, Upsilon_0 absorp1>+  epsilon
#####################################################################################

est_rob_CUADRA <- estimar(y= tecdatos$y,xcenter= tecdatos$x,  ttt= tecdatos$t, 
			ajuste='quadra', freq=freq,  cov_type='gerS',  fLoss='lmrob',
			cterho=3.443689, nresamp=5000, varexp=varexp) 


#######################################
# PERCENTAGE EXPLAINED By 4 directions
#######################################

est_rob_CUADRA$porcentaje
#0.9841754


#############################################
# Store the estimates and compute predictions
# for the residuals
#############################################

beta_rob_CUADRA <- est_rob_CUADRA$beta
gamma_rob_CUADRA <- est_rob_CUADRA$gamma
phies_rob <- est_rob_CUADRA$autofun

coef_rob_CUADRA <-   tecdatos$x %*% phies_rob * dt_tecdatos

predichos_ROB_CUADRA <- est_rob_CUADRA$alfa + coef_rob_CUADRA%*% est_rob_CUADRA$slope_coef + diag(coef_rob_CUADRA%*% est_rob_CUADRA$gama_coef%*% t(coef_rob_CUADRA))


#########################################################
## Residuals from a robust fit
#######################################################
 
residuos_ROB_CUADRA_tecdatos <- tecdatos$y - predichos_ROB_CUADRA

residuos_ROB_LINEAL_tecdatos <- tecdatos$y - predichos_ROB_LINEAL
```


Residuals boxplots both for FLM and FQM are displayed below. 

```{r robust est resbox, echo=FALSE, fig.height=6, fig.width=16}
######################################
# RESIDUAL BOXPLOTS FROM FLM AND FQM
#####################################
par(mfrow=c(1,2))
boxplot(residuos_ROB_LINEAL_tecdatos,col="steelblue", main="FLM")
boxplot(residuos_ROB_CUADRA_tecdatos,col="steelblue", main="FQM")
par(mfrow=c(1,1))
```

In both cases, we show the common outliers between residuals boxplots and functional boxplots.

```{r robust est resboxhide, echo = FALSE, results = FALSE, fig.show='hide'}
##########################################
# Residuals boxplots and atypical data
#########################################
names(residuos_ROB_LINEAL_tecdatos) <- 1:length(residuos_ROB_LINEAL_tecdatos)
atipicos_LINEAL_tecdatos <- as.numeric(names(boxplot(residuos_ROB_LINEAL_tecdatos)$out))
names(residuos_ROB_CUADRA_tecdatos) <- 1:length(residuos_ROB_CUADRA_tecdatos)
atipicos_CUADRA_tecdatos <- as.numeric(names(boxplot(residuos_ROB_CUADRA_tecdatos)$out))
```

```{r robust est resbox id, echo=FALSE}
print("The outliers from the robust linear fit are:")
print(atipicos_LINEAL_tecdatos)
# 7  34  35  43  44  45 118 119 121 122 126 127 128 129 130 140 143 168 171
# 172 185 186 215

print("The common outliers with the functional boxplot are:")
intersect(fd1$outpoint[fd1$outpoint <= length(indices_tecdatos)], atipicos_LINEAL_tecdatos)
 # 35 140

print("The outliers from the robust quadratic fit are:")
print(atipicos_CUADRA_tecdatos)
#  4   8  10  20  31  34  35  38  40  43  44  45 102 108 117 121 122 123 125
# 126 127 129 131 140 171 172 174 175 183 186 211 215

print("The common outliers with the functional boxplot are:")
intersect(fd1$outpoint[fd1$outpoint <= length(indices_tecdatos)], atipicos_CUADRA_tecdatos)
# 35 140
```

Curves corresponding to atypical residuals from FQM are displayed below in red.

```{r robust est atypical curves, echo=FALSE, fig.width=16, fig.height=3}
######################################
# PLOT THE CURVES CORRESPONDING TO 
# ATYPICAL Residuals from FQM
#####################################
matplot(absorp1$argvals, t(absorp1$data), type = "l", lty = 1,lwd=2,
        xlab = "Wavelength", ylab = "X(t)",col="gray60")

matplot(absorp1$argvals, t(absorp1$data[atipicos_CUADRA_tecdatos,]), type = "l", 
	lty = 2,lwd=2,
        xlab = "Wavelength", ylab = "X(t)",col="red",add=TRUE)
```

Beta parameter from FLM fit is displayed below.

```{r robust est beta FLM, echo=FALSE, fig.height=3, fig.width=16}
##############################################
## PLOT BETA FROM FLM FIT
############################################## 
plot(tecdatos$t, beta_rob_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "steelblue", ylab = expression(hat(beta)), main="FLM")
```

Beta and gamma parameters from FQM fit are displayed below.

```{r robust est beta gamma FQM, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## PLOT BETA AND GAMMA FROM FQM FIT
############################################## 
plot(tecdatos$t, beta_rob_CUADRA  , type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "steelblue", ylab = expression(hat(beta)), main="FQM")
 

ene.col=15
colores=rainbow(ene.col, s = 1, v = 1, start = 0, end = max(1, ene.col - 1)/ene.col, alpha = 1)
 
eme= min(gamma_rob_CUADRA)
EME= max(gamma_rob_CUADRA)

#floor(eme)

donde=seq(eme , EME,length=ene.col)

wireframe(gamma_rob_CUADRA,scales = list(arrows = FALSE),   
	xlab="t",ylab="s",zlab=" ",   cex=1.2,row.values=tecdatos$t,
	column.values=tecdatos$t,drape = TRUE,aspect = c(1,1) , 
  	colorkey = TRUE,at=donde, col.regions=colores, main="FQM")
```


Comparative beta from FLM and FQM.

```{r robust est betas, echo=FALSE, fig.height=6, fig.width=16}
par(mfrow=c(1,2))
plot(tecdatos$t, beta_rob_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "steelblue", ylab = expression(hat(beta)), main="FLM")

plot(tecdatos$t, beta_rob_CUADRA  , type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "steelblue", ylab = expression(hat(beta)), main="FQM")
par(mfrow=c(1,1))
```


Residuals vs predicted, both for FLM and FQM are displayed below.

```{r robust est6, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## PLOT residuals vs predicted
############################################## 

par(mfrow=c(1,2))

bb_L=boxplot(residuos_ROB_LINEAL_tecdatos)
bb_C=boxplot(residuos_ROB_CUADRA_tecdatos)

plot( predichos_ROB_LINEAL, residuos_ROB_LINEAL_tecdatos, xlab=expression(hat(y[i])), ylab=expression(r[i]), cex=1.3, cex.lab=1.3, main="FLM")
 abline(h=bb_L$stats[1,1],col="red")
abline(h=bb_L$stats[5,1],col="red")

plot( predichos_ROB_CUADRA, residuos_ROB_CUADRA_tecdatos,
xlab=expression(hat(y[i])), ylab=expression(r[i]), cex=1.3, cex.lab=1.3, main="FQM")
abline(h=bb_C$stats[1,1],col="red")
abline(h=bb_C$stats[5,1],col="red")
par(mfrow=c(1,1))
```


