---
title: "README"
author: "Daniela Laura Parada"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
   md_document:
   variant: markdown_github
---

# Introduction

As it is well known, small proportions of outliers and other atypical observations can affect seriously the estimators for regression models, and the situation in functional linear or quadratic models is not an exception. In fact, different types of outliers may arise when considering functional data: in the functional setting, atypical data might consist of curves that behave differently from the others displaying a persistent behaviour either in shift, amplitude and/or shape making more difficult their detection.

The following is a real data example of the implementation of robust estimators for functional quadratic regression models. This robust proposal involves robust estimators of the principal directions with robust regression estimators based on a bounded loss function and a preliminary residual scale estimator. This is part of a work in progress done in collaboration with Prof. Dra. Graciela Boente [Boente, G. and Parada, D. (2022). _Robust estimation for functional quadratic regression models_. Available at 
https://arxiv.org/abs/2209.02742.]

More precisely, we first compute robust estimators of the principal directions with the aim of providing finite-dimensional candidates for the estimators of both the functional regression parameter and the quadratic operator. We then apply MM--regression estimators (Yohai, 1987) that are based on a bounded loss function and a preliminary residual scale estimator to the residuals obtained from these finite-dimensional spaces. The initial scale estimator ensures that the estimators of the functional regression parameter and the quadratic operator are scale
equivariant, while the bounded loss function and the robust principal directions guarantee that the resulting procedure will be robust against high-leverage outliers. It is worth mentioning that the presence of outliers in the functional covariates may affect the estimation procedure when the sample principal components are used to estimate the regression function and quadratic operator, even when MM-estimators are used. The main reason is that a distorted estimator of the principal direction will affect the scores of all the observations in that direction, that is why robust estimators of the principal direction are needed. We consider the spherical principal components introduced in Locantore et al. (1999) and studied in Gervini (2008), Boente et al. (2014) and Boente et al. (2019).

# Real data example: Tecator

We will use the Tecator data set available in the <code>fda.usc</code> library from <code>R</code> (http://lib.stat.cmu.edu/datasets/tecator). Each one of the 215 observations from this dataset consists of a spectrometric curve that corresponds to the absorbance measured on an equally spaced grid of 100 wavelengths between 850 and 1050 nm with Tecator Infratec Food and Feed Analyzer. The contents of fat protein and moisture were also recorded through analytic chemistry methods. The goal of the analysis is to predict the fat content ($y$) using some characteristics of the spectrometric curve. 

Let's first load some <code>R</code> packages.

```{r installation, results='hide', cache=FALSE, results=FALSE, warning=FALSE, comment=FALSE, message=FALSE}
library('fda')          # fda tools
library('robustbase')   # lmrob 
library(fda.usc)        # tecator dataset and fda tools
library(gdata)          # uppertriangle
library(lattice)        # plot
```

Let's now load some custom <code>R</code> functions that are needed to compute the robust estimators proposed.

```{r installationcustom, results='hide', cache=FALSE, results=FALSE, warning=FALSE, comment=FALSE, message=FALSE}
source('funciones-auxiliares.R') # aux functions
source('descomponer3.R')         # for covariance decomposition
source('minimizar.R')            # minimization menu
```


```{r loading data}
datos <- data(tecator)
absorp <- tecator$absorp.fdata
absorp1 <- fdata.deriv(absorp,nderiv = 1)  #  computes the first derivative
```

Plots of the spectrometric data and it's first derivative are displayed below.

```{r initial plots, echo=FALSE, fig.height=6, fig.width=16}
par(mfrow=c(1,2))
plot(absorp)  #  spectrometric curve
plot(absorp1, main="First derivative")  #  first derivative 
par(mfrow=c(1,1))
```


Functional boxplots of the spectrometric data and it's first derivative are displayed below. In both cases, only few observations are detected as atypical even though is a well-known dataset which atypical data.

```{r functional boxplots, echo=FALSE, fig.height=6, fig.width=16}
par(mfrow=c(1,2))
# First derivative
fd1 <- fbplot(t(absorp1$data), x = absorp1$argvals, method =  "MBD",
       xlab = "Wavelength", ylab = "d(Absorbance, 1)",
       xlim = c(850, 1050), ylim = c(-0.02, 0.05))

# Absorbances
fd0<-fbplot(t(absorp$data), x = absorp$argvals, method = "MBD",
       xlab = "Wavelength", ylab = "Absorbance",
       xlim = c(850, 1050), ylim = c(2, 5.5))
par(mfrow=c(1,1))
```


From now on, we will consider the first derivative of the spectrometric curve, which we denote by $X$, as the functional explanatory variable for the functional regression models to be considered.

```{r X(t), echo=FALSE, fig.width=16, fig.height=6}
par(mar=c(5.1, 24.1, 4.1, 24.1))
matplot(absorp1$argvals, t(absorp1$data), type = "l", lty = 1,
        xlab = "Wavelength (nm)", ylab = "X(t)") 
```


# Robust Estimators under a Functional Linear/Quadratic Model

For the functional regression model, we consider two relationships between the functional explanatory variable $X$ and the scalar response $y$.

Functional Linear Model (FLM): $$y=\alpha+\langle X,\beta\rangle+\epsilon$$

Functional Quadratic Model (FQM): $$y=\alpha+\langle X,\beta\rangle+\langle X,\Upsilon X\rangle+ \epsilon$$

In both cases, the purpose is to estimate the linear coefficient $\beta$ and the quadratic kernel $\upsilon$ related to the $\Upsilon$ operator in the quadratic model (FQM). In that sense, we choose 4 principal directions which explain more than 98% of the total variability as seen below, and provide a robust fit that involves robust estimators of the principal directions with robust regression estimators based on a bounded loss function and a preliminary residual scale estimator.

```{r robust est, echo=FALSE}
set.seed(124) 
covariable <- 'd1' # first derivative  
 
indices <- 1:215  

# indices
tecdatos <- submuestra(indices, covariable = covariable)
 
indices_tecdatos <- indices
dt_tecdatos <- tecdatos$t[2]-tecdatos$t[1]



###############################################
## Choose the number of principal directions
## equal to 4, then varexp should be 1
###############################################


freq=4
varexp=1

est_rob_LINEAL <- estimar(y=tecdatos$y,xcenter= tecdatos$x,  
			ttt= tecdatos$t, ajuste='lineal', freq=freq,  cov_type='gerS',  
			fLoss='lmrob',cterho=3.443689, nresamp=5000, varexp=varexp) 

#######################################
# PERCENTAGE EXPLAINED By 4 directions
#######################################

est_rob_LINEAL$porcentaje

#0.9841754

#############################################
# Store the estimates and compute predictions
# for the residuals
#############################################

beta_rob_LINEAL <- est_rob_LINEAL$beta
 
phies_rob_LINEAL <- est_rob_LINEAL$autofun

coef_rob_LINEAL <-   tecdatos$x %*% phies_rob_LINEAL * dt_tecdatos

predichos_ROB_LINEAL <- est_rob_LINEAL$alfa + coef_rob_LINEAL%*% est_rob_LINEAL$slope_coef  


#####################################################################################
## ROBUST ESTIMATOR UNDER A FQM 
## fat= alfa_0 + < absorp1, beta_0 > + <absorp1, Upsilon_0 absorp1>+  epsilon
#####################################################################################

est_rob_CUADRA <- estimar(y= tecdatos$y,xcenter= tecdatos$x,  ttt= tecdatos$t, 
			ajuste='quadra', freq=freq,  cov_type='gerS',  fLoss='lmrob',
			cterho=3.443689, nresamp=5000, varexp=varexp) 


#######################################
# PERCENTAGE EXPLAINED By 4 directions
#######################################

est_rob_CUADRA$porcentaje
#0.9841754


#############################################
# Store the estimates and compute predictions
# for the residuals
#############################################

beta_rob_CUADRA <- est_rob_CUADRA$beta
gamma_rob_CUADRA <- est_rob_CUADRA$gamma
phies_rob <- est_rob_CUADRA$autofun

coef_rob_CUADRA <-   tecdatos$x %*% phies_rob * dt_tecdatos

predichos_ROB_CUADRA <- est_rob_CUADRA$alfa + coef_rob_CUADRA%*% est_rob_CUADRA$slope_coef + diag(coef_rob_CUADRA%*% est_rob_CUADRA$gama_coef%*% t(coef_rob_CUADRA))


#########################################################
## Residuals from a robust fit
#######################################################
 
residuos_ROB_CUADRA_tecdatos <- tecdatos$y - predichos_ROB_CUADRA

residuos_ROB_LINEAL_tecdatos <- tecdatos$y - predichos_ROB_LINEAL
```


Residuals boxplots both for linear (FLM) and quadratic (FQM) robust fit are displayed below. 

```{r robust est resbox, echo=FALSE, fig.height=6, fig.width=16}
######################################
# RESIDUAL BOXPLOTS FROM FLM AND FQM
#####################################
par(mfrow=c(1,2))
boxplot(residuos_ROB_LINEAL_tecdatos,col="steelblue", main="FLM")
boxplot(residuos_ROB_CUADRA_tecdatos,col="steelblue", main="FQM")
par(mfrow=c(1,1))
```

In both cases, we show the common outliers between residuals boxplots and functional boxplots shown previously. As seen below, only observations 35 and 140 are common outliers with the functional boxplots for both linear (FLM) and quadratic (FQM) robust fit. 

The residual plots show that the functional linear model (FLM) does not seem to provide a reasonable fit for the robust method. Besides, when looking at the residuals from the robust quadratic fit, some atypical residuals are revealed. The boxplot of these residuals allows to identify 32 observations as potential outliers, as seen below.

```{r robust est resboxhide, echo = FALSE, results = FALSE, fig.show='hide'}
##########################################
# Residuals boxplots and atypical data
#########################################
names(residuos_ROB_LINEAL_tecdatos) <- 1:length(residuos_ROB_LINEAL_tecdatos)
atipicos_LINEAL_tecdatos <- as.numeric(names(boxplot(residuos_ROB_LINEAL_tecdatos)$out))
names(residuos_ROB_CUADRA_tecdatos) <- 1:length(residuos_ROB_CUADRA_tecdatos)
atipicos_CUADRA_tecdatos <- as.numeric(names(boxplot(residuos_ROB_CUADRA_tecdatos)$out))
```

```{r robust est resbox id, echo=FALSE}
print("The outliers from the robust FLM fit are:")
print(atipicos_LINEAL_tecdatos)
# 7  34  35  43  44  45 118 119 121 122 126 127 128 129 130 140 143 168 171
# 172 185 186 215

print("The common outliers with the functional boxplot are:")
intersect(fd1$outpoint[fd1$outpoint <= length(indices_tecdatos)], atipicos_LINEAL_tecdatos)
 # 35 140

print("The outliers from the robust FQM fit are:")
print(atipicos_CUADRA_tecdatos)
#  4   8  10  20  31  34  35  38  40  43  44  45 102 108 117 121 122 123 125
# 126 127 129 131 140 171 172 174 175 183 186 211 215

print("The common outliers with the functional boxplot are:")
intersect(fd1$outpoint[fd1$outpoint <= length(indices_tecdatos)], atipicos_CUADRA_tecdatos)
# 35 140
```

The curves corresponding to the 32 atypical observations identified on the residuals boxplot from quadratic robust fit are displayed below in red dashed lines.

```{r robust est atypical curves, echo=FALSE, fig.width=16, fig.height=6}
######################################
# PLOT THE CURVES CORRESPONDING TO 
# ATYPICAL Residuals from FQM
#####################################
par(mar=c(5.1, 24.1, 4.1, 24.1))
matplot(absorp1$argvals, t(absorp1$data), type = "l", lty = 1,lwd=2,
        xlab = "Wavelength", ylab = "X(t)",col="gray60")

matplot(absorp1$argvals, t(absorp1$data[atipicos_CUADRA_tecdatos,]), type = "l", 
	lty = 2,lwd=2,
        xlab = "Wavelength", ylab = "X(t)",col="red",add=TRUE)
```

The linear coefficient estimated from linear (FLM) robust fit, $\hat{\beta}$, is displayed as the red curve below.

```{r robust est beta FLM, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## PLOT BETA FROM FLM FIT
############################################## 
par(mar=c(5.1, 24.1, 4.1, 24.1))
plot(tecdatos$t, beta_rob_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "tomato", ylab = expression(hat(beta)), main="FLM")
```

The estimated parameters $\hat{\beta}$ and $\hat{\upsilon}$ from the quadratic (FQM) robust fit are displayed below as the blue curve and the coloured surface, respectively.

```{r robust est beta gamma FQM, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## PLOT BETA AND GAMMA FROM FQM FIT
############################################## 
par(mar=c(5.1, 24.1, 4.1, 24.1))
plot(tecdatos$t, beta_rob_CUADRA  , type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "steelblue", ylab = expression(hat(beta)), main="FQM")
 

ene.col=15
colores=rainbow(ene.col, s = 1, v = 1, start = 0, end = max(1, ene.col - 1)/ene.col, alpha = 1)
 
eme= min(gamma_rob_CUADRA)
EME= max(gamma_rob_CUADRA)

#floor(eme)

donde=seq(eme , EME,length=ene.col)

wireframe(gamma_rob_CUADRA,scales = list(arrows = FALSE),   
	xlab="t",ylab="s",zlab=" ",   cex=1.2,row.values=tecdatos$t,
	column.values=tecdatos$t,drape = TRUE,aspect = c(1,1) , 
  	colorkey = TRUE,at=donde, col.regions=colores, main="FQM")
```


We can now compare both $\beta$ estimators from linear (FLM) and quadratic (FQM) robust fit. As seen below, the shape is quite similar, but  when quadratic model is considered, the linear estimator take larger absolute values for wavelengths varying between 880 and 980 nm.

```{r robust est betas, echo=FALSE, fig.height=6, fig.width=16}
par(mfrow=c(1,2))
plot(tecdatos$t, beta_rob_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "tomato", ylab = expression(hat(beta)), main="FLM")

plot(tecdatos$t, beta_rob_CUADRA  , type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "steelblue", ylab = expression(hat(beta)), main="FQM")
par(mfrow=c(1,1))
```

```{r robust est betas2, echo=FALSE, fig.height=6, fig.width=16}
par(mar=c(5.1, 24.1, 4.1, 24.1))
plot(tecdatos$t, beta_rob_CUADRA  , type = "l", lwd=3, xlab = "Wavelength", pch = 10,
     col = "steelblue", ylab = expression(hat(beta)))
points(tecdatos$t, beta_rob_LINEAL, type = "l", lwd=3, pch = 10,
     col = "tomato")
legend(1000, -200, legend=c("FLM", "FQM"),
       col=c("tomato", "steelblue"), lty=1, lwd=3, cex=1, bty="n")
```


The residual vs predicted plots, both for linear (FLM) and quadratic (FQM) robust fit are displayed below. These residual plots also show that the functional linear model does not seem to provide a reasonable fit for the robust method. Also, when looking at the residuals from the robust quadratic fit, some atypical residuals are revealed.

```{r robust est6hide, echo=FALSE, fig.show="hide"}
bb_L=boxplot(residuos_ROB_LINEAL_tecdatos)
bb_C=boxplot(residuos_ROB_CUADRA_tecdatos)
```

```{r robust est6, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## PLOT residuals vs predicted
############################################## 
par(mfrow=c(1,2))
plot( predichos_ROB_LINEAL, residuos_ROB_LINEAL_tecdatos, xlab=expression(hat(y[i])), ylab=expression(r[i]), cex=1.3, cex.lab=1.3, main="FLM")
 abline(h=bb_L$stats[1,1],col="red")
abline(h=bb_L$stats[5,1],col="red")

plot( predichos_ROB_CUADRA, residuos_ROB_CUADRA_tecdatos,
xlab=expression(hat(y[i])), ylab=expression(r[i]), cex=1.3, cex.lab=1.3, main="FQM")
abline(h=bb_C$stats[1,1],col="red")
abline(h=bb_C$stats[5,1],col="red")
par(mfrow=c(1,1))
```


# Classic Estimators under a Functional Linear/Quadratic Model and comparison between their robust counterparts

We now want to compare the robust estimators with the ones obtained when using a classical procedure for both linear (FLM) and quadratic (FQM) model. We refer to "classical procedure" as the estimation procedure based on sample principal directions and least squares approach.

As before, we choose 4 principal directions for the classic fit which explain more than 97% of the total variability, as seen below.

```{r classic est, echo=FALSE}
#######################################
## LEAST SQUARE ESTIMATE UNDER FLM
#######################################

est_CL_LINEAL <- estimar(y= tecdatos$y,xcenter= tecdatos$x,  ttt= tecdatos$t, 
			ajuste='lineal', freq=freq,  cov_type='cl',  fLoss='ls',cterho=3.443689, 
			nresamp=5000, varexp=varexp) 



#######################################
# PERCENTAGE EXPLAINED By 4 directions
#######################################

est_CL_LINEAL$porcentaje
#0.9791543


#############################################
# Store the estimates and compute predictions
# for the residuals
#############################################


beta_CL_LINEAL <- est_CL_LINEAL$beta 
phies_CL_LINEAL <- est_CL_LINEAL$autofun

coef_CL_LINEAL <-   tecdatos$x %*% phies_CL_LINEAL * dt_tecdatos

predichos_CL_LINEAL <- est_CL_LINEAL$alfa + coef_CL_LINEAL%*% est_CL_LINEAL$slope_coef 


#######################################
## Estimador  CLASICO  CUADRATICO
#######################################

est_CL_CUADRA <- estimar(y= tecdatos$y,xcenter= tecdatos$x,  ttt= tecdatos$t, 
			ajuste='quadra', freq=freq,  cov_type='cl',  
			fLoss='ls',cterho=3.443689, nresamp=5000, varexp=varexp) 


#############################################
# Store the estimates and compute predictions
# for the residuals
#############################################


beta_CL_CUADRA <- est_CL_CUADRA$beta
gamma_CL_CUADRA <- est_CL_CUADRA$gamma
phies_CL <- est_CL_CUADRA$autofun
 
coef_CL_CUADRA <-   tecdatos$x %*% phies_CL * dt_tecdatos

predichos_CL_CUADRA <- est_CL_CUADRA$alfa + coef_CL_CUADRA%*% est_CL_CUADRA$slope_coef + diag(coef_CL_CUADRA%*% est_CL_CUADRA$gama_coef%*% t(coef_CL_CUADRA))



#############################################
#  RESIDUALS FROM THE LEAST SQUARES FIT
#############################################
 
 
residuos_CL_CUADRA_tecdatos <- tecdatos$y - predichos_CL_CUADRA

residuos_CL_LINEAL_tecdatos <- tecdatos$y - predichos_CL_LINEAL


```

The linear coefficient estimated from linear (FLM) classic fit, $\hat{\beta}$, is displayed as the red curve below.

```{r classic est beta FLM, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## PLOT BETA FROM FLM FIT
############################################## 
par(mar=c(5.1, 24.1, 4.1, 24.1))
plot(tecdatos$t, beta_CL_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "red", ylab = expression(hat(beta)),main="FLM",
	 ylim=c(min(c(beta_CL_LINEAL,beta_rob_LINEAL)), max(c(beta_CL_LINEAL,beta_rob_LINEAL))))

```

We can now compare $\beta$ estimates in the linear model (FLM) for both classic (CL) and robust (ROB) fit. As seen beofre, the shape is quite similar, but when robust fit is considered, the linear estimator take larger absolute values for wavelengths varying between 910 and 960 nm.


```{r robvsclas est beta FLM, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## COMPARE WITH THE ROBUST ONE
############################################## 
par(mar=c(5.1, 24.1, 4.1, 24.1))
plot(tecdatos$t, beta_CL_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "red", ylab = expression(hat(beta)), main="FLM",
	 ylim=c(min(c(beta_CL_LINEAL,beta_rob_LINEAL)), max(c(beta_CL_LINEAL,beta_rob_LINEAL))))
 

lines(tecdatos$t, beta_rob_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "blue")

legend(1000, -100, legend=c("CL","ROB"),
       col=c("red", "blue"), lty=1, lwd=3, cex=1, bty="n")

```


The estimated parameters $\hat{\beta}$ and $\hat{\upsilon}$ from the quadratic (FQM) classic fit are displayed below as the red curve and the coloured surface, respectively.

```{r classic est beta gamma FQM, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## PLOT BETA AND GAMMA FROM FQM FIT
############################################## 
par(mar=c(5.1, 24.1, 4.1, 24.1))
plot(tecdatos$t, beta_CL_CUADRA, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "red", ylab = expression(hat(beta)),main="FQM",
     ylim=c(min(c(beta_CL_CUADRA,beta_rob_CUADRA)), max(c(beta_CL_CUADRA,beta_rob_CUADRA))))

ene.col=15
colores=rainbow(ene.col, s = 1, v = 1, start = 0, end = max(1, ene.col - 1)/ene.col, alpha = 1)
 
eme= min(gamma_CL_CUADRA)
eMe= max(gamma_CL_CUADRA)

#floor(eme)

donde=seq(eme ,eMe  ,length=ene.col)
 
wireframe(gamma_CL_CUADRA,scales = list(arrows = FALSE),   
		xlab="t",ylab="s",zlab=" ",   cex=1.2,
		row.values=tecdatos$t,column.values=tecdatos$t,drape = TRUE,
  		colorkey = TRUE,at=donde, col.regions=colores, main="FQM")
```

We can now compare $\beta$ estimates in the quadratic model (FQM) between classic (CL) and robust (ROB) fit. As seen below, the shape is quite similar, but when robust fit is considered, the linear estimator take larger absolute values for wavelengths varying between 900 and 980 nm.

```{r robvsclas est beta FQM, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## COMPARE WITH THE ROBUST ONE
############################################## 
par(mar=c(5.1, 24.1, 4.1, 24.1))
plot(tecdatos$t, beta_CL_CUADRA, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "tomato", ylab = expression(hat(beta)), main="FQM",
     ylim=c(min(c(beta_CL_CUADRA,beta_rob_CUADRA)), max(c(beta_CL_CUADRA,beta_rob_CUADRA))))
 

lines(tecdatos$t, beta_rob_CUADRA, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "steelblue")

legend(1000, -200, legend=c("CL","ROB"),
       col=c("tomato", "steelblue"), lty=1, lwd=3, cex=1, bty="n")

```
Residual vs predicted plots, both for linear (FLM) and quadratic (FQM) model in the classic fit are displayed below. These residual plots show, again, that the functional linear model (FLM) does not seem to provide a reasonable fit for the classical method either.


```{r classic est6, echo=FALSE, fig.height=6, fig.width=16}
##############################################
## PLOT residuals vs predicted
############################################## 
par(mfrow=c(1,2))
plot( predichos_CL_LINEAL,residuos_CL_LINEAL_tecdatos, xlab=expression(hat(y[i])), ylab=expression(r[i]), cex=1.3, cex.lab=1.3,
      main="FLM")

plot( predichos_CL_CUADRA, residuos_CL_CUADRA_tecdatos,
xlab=expression(hat(y[i])), ylab=expression(r[i]), cex=1.3, cex.lab=1.3, main="FQM")
par(mfrow=c(1,1))
```

# Analysis without outliers

We now repeat the classical analysis, both for linear (FLM) and quadratic (FQM) model, computed without the detected potential outliers that have been identified in the residual boxplot from the robust quadratic fit(these 32 observationes were manually removed). We will refer it as "CL without out". Again, we choose 4 principal directions which explain more than 97% of the variability as seen below. 

```{r echo=FALSE}
############################ 
## Data without outliers
############################

tecdatos_sin_out <- submuestra(indices_tecdatos[-atipicos_CUADRA_tecdatos], covariable = covariable)


#######################################
## CLASSICAL ESTIMATOR WITHOUT OULTIERS
## FLM
#######################################

est_CL_LINEAL_sin_out <- estimar(y= tecdatos_sin_out$y,xcenter= tecdatos_sin_out$x,  ttt= tecdatos_sin_out$t, 
				ajuste='lineal', freq=freq,  cov_type='cl',  fLoss='ls',cterho=3.443689, 
				nresamp=5000, varexp=varexp) 

est_CL_LINEAL_sin_out$porcentaje
#0.9774696

#####################################################
# Store parameters and compute predicted values
######################################################


beta_CL_LINEAL_sin_out <- est_CL_LINEAL_sin_out$beta 
phies_CL_LINEAL_sin_out <- est_CL_LINEAL_sin_out$autofun

coef_CL_LINEAL_sin_out <-   tecdatos_sin_out$x %*% phies_CL_LINEAL_sin_out * dt_tecdatos

predichos_CL_LINEAL_sin_out <- est_CL_LINEAL_sin_out$alfa + coef_CL_LINEAL_sin_out%*% est_CL_LINEAL_sin_out$slope_coef 


######################################
## CLASSICAL ESTIMATOR WITHOUT OULTIERS
## FQM
#######################################

est_CL_CUADRA_sin_out <- estimar(y= tecdatos_sin_out$y,xcenter= tecdatos_sin_out$x,  
				ttt= tecdatos_sin_out$t, 
				ajuste='quadra', freq=freq,  cov_type='cl',  
				fLoss='ls',cterho=3.443689, nresamp=5000, varexp=varexp) 

######################################################
# Store parameters and compute predicted values
######################################################

beta_CL_CUADRA_sin_out <- est_CL_CUADRA_sin_out$beta
gamma_CL_CUADRA_sin_out <- est_CL_CUADRA_sin_out$gamma
phies_CL_sin_out <- est_CL_CUADRA_sin_out$autofun
 
coef_CL_CUADRA_sin_out <-   tecdatos_sin_out$x %*% phies_CL_sin_out * dt_tecdatos

predichos_CL_CUADRA_sin_out <- est_CL_CUADRA_sin_out$alfa + coef_CL_CUADRA_sin_out%*% est_CL_CUADRA_sin_out$slope_coef + 
diag(coef_CL_CUADRA_sin_out%*% est_CL_CUADRA_sin_out$gama_coef%*% t(coef_CL_CUADRA_sin_out))


######################################################
## RESIDUALS FROM THE CLASSICAL FIT WITHOUT OUTLIERS
###################################################### 

residuos_CL_CUADRA_tecdatos_sin_out <- tecdatos_sin_out$y - predichos_CL_CUADRA_sin_out

residuos_CL_LINEAL_tecdatos_sin_out <- tecdatos_sin_out$y - predichos_CL_LINEAL_sin_out

```

Predicted vs residuals plots in the linear (FLM) and quadratic (FQM) classic without outliers (CL without out) fit are shown in the first pair of plots. 

In the second pair of plots, we show estimated $\beta$ parameters for classic (CL), robust (ROB) and classic after outliers removal (CL without out) fit, both for linear (FLM) and quadratic (FQM) model. As seen below, when estimating the linear regression function $\beta$, the classical estimators computed without the detected potential outliers are very close to the robust ones, that is, the robust estimator behaves similarly to the classical one if one were able to manually remove suspected outliers.

```{r plotscompar, echo=FALSE, fig.height=6, fig.width=16}
par(mfrow=c(1,2))

plot( predichos_CL_LINEAL_sin_out,residuos_CL_LINEAL_tecdatos_sin_out, xlab=expression(hat(y[i])), ylab=expression(r[i]), cex=1.3, cex.lab=1.3, main="FLM without out")
plot( predichos_CL_CUADRA_sin_out, residuos_CL_CUADRA_tecdatos_sin_out,
xlab=expression(hat(y[i])), ylab=expression(r[i]), cex=1.3, cex.lab=1.3,
main="FQM without out")


plot(tecdatos$t, beta_CL_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "red", ylab = expression(hat(beta)),main="FLM",
	 ylim=c(min(c(beta_CL_LINEAL,beta_rob_LINEAL)), max(c(beta_CL_LINEAL,beta_rob_LINEAL))))
 
lines(tecdatos$t, beta_rob_LINEAL, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "blue")

lines(tecdatos$t, beta_CL_LINEAL_sin_out, type = "l", lty=2,lwd=4,xlab = "Wavelength", pch = 10,
     col = "pink")
legend(1000, -100, legend=c("CL", "ROB", "CL without out"),
       col=c("red", "blue", "pink"), lty=c(1,1,2), lwd=3, cex=1, bty="n")


plot(tecdatos$t, beta_CL_CUADRA, type = "l",  lwd=3,xlab = "Wavelength", pch = 10,
     col = "tomato", ylab = expression(hat(beta)),main="FQM",
     ylim=c(min(c(beta_CL_CUADRA,beta_rob_CUADRA)), max(c(beta_CL_CUADRA,beta_rob_CUADRA))))
 
lines(tecdatos$t, beta_rob_CUADRA, type = "l", lwd=3,xlab = "Wavelength", pch = 10,
     col = "blue")
 
lines(tecdatos$t, beta_CL_CUADRA_sin_out, type = "l", lty=2,lwd=4,xlab = "Wavelength", pch = 10,
     col = "pink")
legend(1000, -200, legend=c("CL", "ROB", "CL without out"),
       col=c("tomato", "blue", "pink"), lty=c(1,1,2), lwd=3, cex=1, bty="n")

par(mfrow=c(1,1))
```

Note that for both the linear coefficient $\beta$ and the quadratic kernel $\upsilon$, the shape of the classical estimators computed without the suspected atypical observations, resembles that of the robust ones. To visualize more clearly the similarity between the surfaces related to the quadratic kernel estimators, we present the surfaces differences between classic (CL) and robust (ROB) fit, and between classic without outliers (CL without out) and robust (ROB) fit, both for quadratic model. As seen below, when estimating the quadratic operator, the classical estimators computed without the detected potential outliers are very close to the robust ones, that is, the robust estimator behaves similarly to the classical one if one were able to manually remove suspected outliers.

```{r dif suf CL ROB, echo=FALSE, fig.height=6, fig.width=16}
par(mar=c(5.1, 24.1, 4.1, 24.1))


DIF_CL_ROB <- gamma_CL_CUADRA - gamma_rob_CUADRA
DIF_CL_sin_out_ROB <- gamma_CL_CUADRA_sin_out- gamma_rob_CUADRA

eme_D=min(c(DIF_CL_ROB,DIF_CL_sin_out_ROB))
EME_D=max(c(DIF_CL_ROB,DIF_CL_sin_out_ROB))



donde_D=seq(eme_D ,EME_D  ,length=ene.col)

wireframe(DIF_CL_ROB ,scales = list(arrows = FALSE),   
	xlab="t",ylab="s",zlab=" ",   cex=1.2,zlim=c(eme_D ,EME_D),
	row.values=tecdatos$t,column.values=tecdatos$t,drape = TRUE,
  	colorkey = TRUE,aspect = c(1,1),screen=list(z= 30,x=- 70),
 	at=donde_D, col.regions=colores, main="Difference between estimated surfaces: CL and ROB ") 


wireframe(DIF_CL_sin_out_ROB ,scales = list(arrows = FALSE),   
	xlab="t",ylab="s",zlab=" ",   cex=1.2,zlim=c(eme_D ,EME_D),
	row.values=tecdatos$t,column.values=tecdatos$t,drape = TRUE,
  	colorkey = TRUE, aspect = c(1,1),screen=list(z= 30,x=- 70),
	at=donde_D, col.regions=colores, main="Difference between estimated surfaces: CL without out and ROB") 


eme1=min(DIF_CL_sin_out_ROB)
EME1= max(DIF_CL_sin_out_ROB)


donde1=seq(eme1 ,EME1  ,length=ene.col)

wireframe(DIF_CL_sin_out_ROB ,scales = list(arrows = FALSE),   
	xlab="t",ylab="s",zlab=" ",   cex=1.2,zlim=c(eme1 ,EME1),
	row.values=tecdatos$t,column.values=tecdatos$t,drape = TRUE,
  	colorkey = TRUE, aspect = c(1,1),screen=list(z= 30,x=- 70),
	at=donde1, col.regions=colores, main="Difference between estimated surfaces: CL without out and ROB (zoomed)") 
 
```

